\Title

### 2013-05-31 synmut.tex
Implemented Richard's suggestions (including 'mutation' -> SNV)

\begin{abstract}
\noindent
Intrapatient HIV-1 evolution is dominated by selection on the protein level in the
arms race with the adaptive immune system. When cytotoxic CD8${}^+$ T-cells or 
neutralizing antibodies
target a new epitope, the virus often escapes via nonsynonymous mutations that
impair recognition. Synonymous mutations do not affect this interplay and are
often assumed to be neutral.
We analyze longitudinal intrapatient data from the \shankaregion{} part of the
envelope gene (\env{}) and observe that synonymous derived alleles rarely
fix even though they often reach high frequencies in the viral population.
We find that synonymous mutations that disrupt base pairs in RNA stems flanking
the variable loops of gp120 are more likely to be lost than other synonymous
changes, hinting at a direct fitness effect of these RNA hairpins in HIV-1.
Computational modeling indicates that these synonymous mutations have a
selection coefficient of the order of $-0.002$ and that they are
brought up to high frequency by hitchhiking on neighboring beneficial
nonsynonymous alleles.
The patterns of fixation of nonsynonymous mutations estimated from the
longitudinal data and comparisons with computer models
suggest that escape mutations in \shankaregion{} are only transiently
beneficial, either because the immune system is catching up or because of
competition between equivalent escapes.

\end{abstract}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\maketitle
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\section{Introduction}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
HIV-1 evolves rapidly within a single host during the course of the infection.
This evolution is driven by strong selection imposed by the host immune system
via cytotoxic CD8${}^+$ T-cells (CTLs) and neutralizing antibodies
(nAbs)~\citep{rambaut_causes_2004} and facilitated by the high mutation rate
~\citep{mansky_lower_1995,abram_nature_2010}. When the host develops a CTL or
nAb response against a particular HIV-1 epitope, mutations in the viral genome that
reduce or prevent recognition of the epitope frequently emerge. Escape mutations
in epitopes targeted by CTLs typically evolve during early infection and spread
rapidly through the population~\citep{mcmichael_immune_2009}. During chronic
infection, the most rapidly evolving parts of the HIV-1 genome are the variable
loops V1-V5 in the envelope protein gp120, which change to avoid recognition by
nAbs. Escape mutations in \env, the gene encoding gp120, spread through the
viral population within a few months.
Consistent with this time scale, it is found that serum from a particular time
typically neutralizes virus extracted more than 3-6 months earlier but not contemporary
virus \citep{richman_rapid_2003}.

Escape mutations are selected because they change the amino acid sequence of
viral proteins in a way that reduces antibody binding or epitope presentation.
Conversely, synonymous mutations do not modify the viral protein and are
commonly used as approximately neutral markers in studies of viral evolution.
Neutral markers are very useful since their dynamics can be compared to that of
putatively functional sites to detect purifying or directional selection
\citep{Bhatt:2011p43255,Hurst:2002p32608,Chen:2004p22606}. In addition to
maintaining protein function and avoiding the adaptive immune recognition,
however, the HIV-1 genome has to ensure efficient processing and translation,
nuclear export, and packaging into the viral capsid: all these processes operate
at the RNA level and are sensitive to synonymous changes. For example, the HIV-1
\rev{} response element (RRE) in \env{} enhances nuclear export of full length
or partially spliced viral transcripts via a complex hairpin RNA
structure~\citep{fernandes_hiv-1_2012}. Another well studied case is the
interaction between viral reverse transcriptase, viral ssRNA, and the host
tRNA$^\text{Lys3}$: the latter is required for priming reverse transcription
(RT) and is bound by a pseudoknotted RNA structure in the viral 5' untranslated
region~\citep{barat_interaction_1991, paillart_vitro_2002}.

@@

In this paper, we characterize the dynamics of synonymous mutations in \env{}
and show that, in the region of the V loops, a large fraction of these mutations
is deleterious. We argue that deleterious synonymous mutations
spread in the viral population via genetic hitchhiking due to limited recombination in HIV-1
populations~\citep{neher_recombination_2010, batorsky_estimate_2011}. We show a
strong correlation between the fate of a synonymous mutation and the surrounding RNA
structure, suggesting an important role of RNA hairpins around V-loops
in HIV-1 evolution. We then compare our observations to computational models and derive
estimates for the effect of synonymous mutations on viral fitness. Extending the
analysis of fixation probabilities to the nonsynonymous mutations, we show that
time-dependent selection or strong competition of escape mutations inside the
same epitope are necessary to explain the observed patterns of fixation and
loss.

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\section{Results}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
The central quantity we investigate is the probability of fixation of a
mutation, conditional on its population frequency. A neutral mutation
segregating at frequency $\nu$ has a probability $\pfix(\nu) = \nu$ to
spread through the population and fix; in the rest of the cases, i.e. with
probability $1-\nu$, it goes extinct. For instance, if a mutation is observed in
half of the sequences, it will fix with a probability of 50\% (see inset in
figure \ref{fig:aftsyn}).
%This is a  consequence of the fact that (i) exactly one
%individual in the current population will be the common ancestor of the entire
%future population at a particular site and (ii) this ancestor has a probability
%$\nu$ of carrying the mutation (assuming the neutral mutation is not
%preferentially associated with genomes of high or low fitness).
Deleterious or
beneficial mutations fix less or more often than neutral ones, respectively.
\FIG{aft} shows the time course of the frequencies of all synonymous and
nonsynonymous viral mutations observed in \env, in \shankaregion, in a chronically
HIV-1 infected patient (p10 from\citet{shankarappa_consistent_1999}). Despite many synonymous
mutations reaching high frequency, few fix (panel~\ref{fig:aftsyn}); in
constrast, many nonsynonymous mutations fix (panel~\ref{fig:aftnonsyn}).
Strictly speaking, no mutation in the HIV-1 population ever fixes because the 
mutation rate and the population size are large. Therefore, we define ``fixation'' or ``loss''
by not observing the mutation in the sample.

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\subsection{Many synonymous polymorphisms in \shankaregion{} are deleterious}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
We study the dynamics and fate of synonymous mutations more quantitatively by
analyzing data from seven patients from
\citet{shankarappa_consistent_1999} and \citep{liu_selection_2006} as well as three patients from
\citet{bunnik_autologous_2008} (patients with strong viral population structure
were excluded from the analysis, because it is impossible to measure
frequencies; see methods and \figurename~S\PCApat). The
former data set is restricted to the \shankaregion{} region of \env, while
the data from \citet{bunnik_autologous_2008} cover most of \env.
We consider all mutations in a frequency interval $[\nu_0-\delta\nu, \nu_0+\delta\nu]$ at some time
$t$, and calculate the fraction that are still observed at later times $t+\Delta t$. Plotting this fraction against the time interval $\Delta t$, we see that
most synonymous mutations segregate for roughly one year and are lost much more
frequently than expected (panel \ref{fig:fixp1}). The long-time probability of
fixation, $\pfix$, is shown as a function of the
initial frequency $\nu_0$ in panel~\ref{fig:fixp2}. 
We find that $\pfix$ of synonymous mutations is far below
the neutral expectation in  \shankaregion (red line).  Outside of \shankaregion, using data from
\citet{bunnik_autologous_2008} only, we find no such reduction in $\pfix$ (green
line).
Restricted to the \shankaregion{} region, the sequence samples from
\citet{bunnik_autologous_2008} are fully compatible with data from
\citet{shankarappa_consistent_1999}. The nonsynonymous mutations seem to follow
more or less the neutral expectation (blue line) -- a point to which we will
come back below.

\begin{figure}
\begin{center}
\subfloat{\includegraphics[width=\linewidth]
{Shankarappa_allele_freqs_trajectories_syn_p10.pdf}
\label{fig:aftsyn}}\\
 \subfloat{\includegraphics[width=\linewidth]
{Shankarappa_allele_freqs_trajectories_nonsyn_p10.pdf}
\label{fig:aftnonsyn}}
\caption{Time series of frequencies
of synonymous (A) and nonsynonymous (B) derived alleles in \env, 
\shankaregion, from patient 10~\cite{shankarappa_consistent_1999}.
While many nonsynonymous mutations  fix, few synonymous
mutations do even though they are frequently observed at intermediate
frequencies. Colors indicate the position of the site along the \shankaregion{} region
(blue to red). Inset: the fixation probability $\pfix$ of a neutral
mutation that reached 50\% frequency is one half.}
\label{fig:aft}
\end{center}
\end{figure}

When interpreting these results for the fixation probabilities, it is important
to distinguish between random mutations and polymorphisms observed at a certain
frequency since the latter have already been filtered by selection.
A polymorphism could be beneficial to the virus and on its way to fixation. In
this case, we expect that it fixes almost surely given that we see it at high
frequency. If, on the other hand, the polymorphism is deleterious it must have
reached a high frequency by chance (genetic drift or hitchhiking), and
we expect that selection drives it out of the population again. Hence our
observations suggest that many of the synonymous polymorphisms at intermediate
frequencies in the part of \env{} that includes \shankaregion{} are
deleterious, while outside this region most polymorphisms are roughly
neutral. Note that this does not imply that all synonymous mutations outside 
\shankaregion{} are neutral -- only those mutations observed at high frequencies, which
have experienced selection for some time, tend to be neutral.

\begin{figure}
\begin{center}
\subfloat{\includegraphics[width=0.9\linewidth]{fixation_times.pdf}
\label{fig:fixp1}}\\
\subfloat{\includegraphics[width=0.9\linewidth]{fixation_probabilities.pdf}
\label{fig:fixp2}}
\caption{Fixation and loss of synonymous mutations.
Panel A) shows how quickly synonymous mutations are purged from the populations. 
Specifically, the figure shows the fraction of mutations that are still observed
after $\Delta t$ days, conditional on being observed in one of the three frequency 
intervals (different colors). 
In each frequency interval, the fraction of synonymous
mutations that ultimately survive is the fixation probability $\pfix$ conditional on the
initial frequency. The neutral expectation for $\pfix=\nu_0$ is indicated by 
dashed horizontal lines.
Panel B) shows the fixation probability of derived synonymous
alleles as a function of $\nu_0$. Polymorphisms within \shankaregion{} fix less
often than expected for neutral mutations (indicated by the diagonal line).
This suppression is not observed in other parts of \env~or for nonsynonymous mutations.
The horizontal error bars on the abscissa are bin sizes, the vertical ones the
standard deviation after 100 patient bootstraps of the data. Data from
refs.~\cite{shankarappa_consistent_1999,liu_selection_2006, bunnik_autologous_2008}.}
\label{fig:fixp}
\end{center}
\end{figure}

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\subsection{Synonymous mutations in \shankaregion{} tend to disrupt RNA stems}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
One possible explanation for lack of fixation of synonymous mutations in
\shankaregion{} is secondary structures in the viral RNA, the disruption of which
is deleterious to the virus \citep{forsdyke_reciprocal_1995,
snoeck_mapping_2011, sanjuan_interplay_2011}.

The propensity of nucleotides in the HIV-1 genome to form base pairs has been
measured using the SHAPE assay, a biochemical reaction preferentially altering
unpaired bases (the HIV-1 genome is a single stranded RNA)
\citep{watts_architecture_2009}. The SHAPE assay has shown that the variable
regions V1-V5 tend to be unpaired, while the conserved regions between those
variable regions form stems.  We aligned the within-patient sequence samples 
to the reference NL4-3 strain used in \citet{watts_architecture_2009} and 
assigned SHAPE reactivities to most positions in the alignment. 
We then calculated the distributions of SHAPE reactivities for synonymous 
polymorphisms that fixed or were subsequently lost (only polymorphisms with 
frequencies above 15\%).
As shown in \FIG{SHAPEA}, the reactivities of fixed alleles (red
histogram) are systematically larger than those of alleles that are lost (blue)
(Kolmogorov-Smirnov test on the cumulative distribution, $p\approx 0.002$). In
other words, alleles that are likely to break RNA helices are also more likely
to revert and finally be lost from the population (restoring the helix). The average over all
mutations that are not observed (green) lies between those that fix and
those that get lost. Note that this analysis will be sensitive only at positions
where the base pairing pattern of NL4-3 agrees with that of each patient's
initial consensus sequence (it is thus statistically conservative).

To test the hypothesis that mutations in \shankaregion{} are lost because they
break stems in the conserved stretches between the V loops, we consider
mutations in V loops and conserved parts separately. The greatest
depression in fixation probability is observed in the conserved stems, while the
V loops show little deviation from the neutral signature, see
\FIG{SHAPEB}. This is suggestive of important RNA helices in conserved
regions between loops.

In addition to RNA secondary structure, we have considered other possible
explanations for a fitness cost of some synonymous mutations, in particular codon
usage bias (CUB). HIV-1 is known to prefer A-rich codons over highly expressed
human codons~\citep{jenkins_extent_2003,kuyl_biased_2012}. We do not find,
however, any evidence for a contribution of average CUB to the ultimate fate of
synonymous alleles; consistently, HIV-1 does not seem to adapt its codon usage to
its human host cells at the macroevolutionary level \citep{kuyl_biased_2012}.

\begin{figure}
\begin{center}
\subfloat{\includegraphics[height=0.50\linewidth]{reactivities_histograms_syn.pdf}\label{fig:SHAPEA}}
\subfloat{\includegraphics[height=0.49\linewidth]{fixation_probabilities_VnonV.pdf}\label{fig:SHAPEB}}
\caption{Permissible synonymous mutations tend to be unpaired.
Panel A) shows the distribution of SHAPE reactivities among sites at which synonymous 
mutations fixed (red), sites at which mutations reached frequencies above 15\% but
were susequently lost (blue), and sites at which no mutations were observed (green) 
(all categories are restricted to the regions V1-V5$\pm 100$bp).
Sites at which mutations fixed tend to have higher SHAPE reactivities, corresponding to
less base pairing, than those at which mutations are lost.
Sites at which no mutations are observed show an intermediate distribution of SHAPE values.
Panel B) shows the fixation probability of synonymous mutations in
\shankaregion{} separately for variable regions V3-V5 and the connecting conserved 
regions C2-C4 that harbor RNA stems. As expected, the fixation probability is lower
inside the conserved regions. Data from Refs.~\cite{shankarappa_consistent_1999,
bunnik_autologous_2008, liu_selection_2006}.}
\label{fig:SHAPE}
\end{center}
\end{figure}

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\subsection{Deleterious mutations are brought to high frequency by hitchhiking}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

While the observation that some fraction of synonymous mutations is deleterious
is not unexpected, it seems odd that we observe them at high population
frequency and that the fixation probability is reduced only in parts of the
genome (in \shankaregion{} but not in the rest of \env{}; compare the red
triangle line versus the green square line in \FIG{fixp2}).
The region \shankaregion{} undergoes frequent adaptive changes to evade
recognition by neutralizing antibodies \cite{williamson_adaptation_2003,
richman_rapid_2003}. Due to the limited amount of recombination in HIV-1
\cite{neher_recombination_2010, batorsky_estimate_2011}, deleterious mutations
that are linked to adaptive variants can reach high frequency. This process is
known as hitchhiking \citep{smith_hitch-hiking_1974} or genetic draft
\citep{gillespie_genetic_2000,neher_genetic_2011}. Hitchhiking is  apparent in
\FIG{aft}, which shows that many mutations change rapidly in frequency as a
flock.

The approximate magnitude of the deleterious effects can be estimated from
\FIG{fixp1}, which shows the distribution of times after which synonymous
alleles at intermediate frequencies become fixed or lost. The typical time to
loss is of the order of 500 days. If this loss is driven by the deleterious
effect of the mutation, this corresponds to deleterious effects $s_d$ of the
order of $- 0.002$ per day. (This is only an average estimate: every single
mutation is expected to have a slightly different fitness effect.)

>>>>

\begin{abstract}
\noindent
Intrapatient HIV-1 evolution is dominated by selection on the protein level in
the arms race with the adaptive immune system. When cytotoxic CD8${}^+$ T-cells
or neutralizing antibodies target a new epitope, the virus often escapes via
nonsynonymous mutations that impair recognition. Synonymous mutations do not
affect this interplay and are often assumed to be neutral. We analyze
longitudinal intrapatient data from the \shankaregion{} part of the envelope
gene (\env{}) and observe that synonymous mutations rarely spread even though
they often reach high frequencies in the viral population. Using published data
from the SHAPE assay, we find that synonymous mutations that disrupt base pairs
in RNA stems flanking the V loops of gp120 are more likely to be lost than other
synonymous changes, suggesting a function of these RNA hairpins in HIV-1.
Computational modeling indicates that these synonymous mutations have a
selection coefficient of the order of $-0.002$ and that they are brought up to
high frequency by genetic hitchhiking on neighboring beneficial variants. We
conclude that, contrary to common assumptions, synonymous mutations in the V
loops region are neither independent of the escape patterns nor neutral; further
studies are needed to clarify the function of the RNA hairpins found there.
\end{abstract}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\maketitle
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\section{Introduction}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
HIV-1 evolves rapidly within a single host during the course of the infection.
This evolution is driven by strong selection imposed by the host immune system
via cytotoxic CD8${}^+$ T-cells (CTLs) and neutralizing antibodies
(nAbs)~\citep{rambaut_causes_2004} and facilitated by the high mutation rate
~\citep{mansky_lower_1995,abram_nature_2010}. When the host develops a CTL or
nAb response against a particular HIV-1 epitope, mutations in the viral genome that
reduce or prevent recognition of the epitope frequently emerge. Escape mutations
in epitopes targeted by CTLs typically emerge during early infection and spread
rapidly through the population~\citep{mcmichael_immune_2009}. During chronic
infection, the most rapidly evolving parts of the HIV-1 genome are the variable
loops V1-V5 in the envelope protein gp120, which change to avoid recognition by
nAbs. Escape mutations in \env, the gene encoding gp120, spread through the
viral population within a few months.
Consistent with this time scale, it is found that serum from a particular time
typically neutralizes virus extracted more than 3-6 months earlier but not contemporary
virus \citep{richman_rapid_2003}.

Escape mutations are selected because they change the amino acid sequence of
viral proteins in a way that reduces antibody binding or epitope presentation.
Conversely, synonymous mutations do not modify the viral protein and are
commonly used as approximately neutral markers in studies of viral evolution.
Neutral markers are very useful as a negative control for detecting selected
sites \citep{Bhatt:2011p43255,Hurst:2002p32608,Chen:2004p22606}. In addition to
maintaining protein function and avoiding the adaptive immune recognition,
however, the HIV-1 genome has to ensure efficient processing and translation,
nuclear export, and packaging into the viral capsid: all these processes operate
at the RNA level and are sensitive to synonymous changes. For example, the HIV-1
\rev{} response element (RRE) in \env{} enhances nuclear export of full length
or partially spliced viral transcripts via a complex hairpin RNA structure
\citep{fernandes_hiv-1_2012}. Another well studied case is the interaction
between viral reverse transcriptase, viral ssRNA, and the host
tRNA$^\text{Lys3}$: the latter is required for priming reverse transcription
(RT) and is bound by a pseudoknotted RNA structure in the viral 5' untranslated
region~\citep{barat_interaction_1991, paillart_vitro_2002}. These RNA
structures, but not the exact base pairs, are well conserved between HIV-1 and
SIV \citep{pollom_comparison_2013}.

@@

In this paper, we characterize the dynamics of synonymous mutations in \env{}
and show that, in the region of the V loops, a large fraction of these mutations
is deleterious. Contrary to common assumptions, deleterious synonymous mutations
rise in frequency in the viral population via genetic hitchhiking due to limited
recombination in HIV-1 populations~\citep{neher_recombination_2010,
batorsky_estimate_2011}. We show a strong correlation between the fate of a
synonymous mutation and the surrounding RNA structure, suggesting an important
role of RNA hairpins around V-loops in HIV-1 evolution. We then compare our
observations to computational models and derive estimates for the effect of
synonymous mutations on viral fitness. Extending the analysis of fixation
probabilities to the nonsynonymous mutations, we show that time-dependent
selection or strong competition of escape mutations inside the same epitope are
necessary to explain the observed patterns of fixation and loss.

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\section{Results}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
In HIV-1, because of the high mutation rate, single nucleotide variants (SNVs)
are continuously injected into the viral population. We say that an SNV has a
frequency $\nu$ at a certain time if it is found in a fraction $\nu$, between
zero and one, of the sequences from that time point. An SNV rises or falls in
frequency by virtue of both (i) its own effect on fitness and escape and (ii)
the effect of its genetic background. When an SNVs is present in all sequences
at a certain time point, we say it has ``fixed''; when it is completely absent,
we say it was ``lost'' or is ``extinct''. Eventually, all SNVs will fix or be lost; the
probability of fixation is higher for beneficial SNVs, and lower for deleterious
ones. The intermediate case are neutral SNVs: a neutral mutation segregating at
frequency $\nu$ has a probability $\pfix(\nu) = \nu$ to spread through the
population and fix; in the rest of the cases it goes extinct. For instance, if
a neutral SNV is observed in half of the sequences, it will fix with a probability of
50\% (see inset in figure \ref{fig:aftsyn}).

The problem with the classification based on beneficial/neutral/deleterious is
that, given an SNV, we do not know its class \textit{a priori}. Synonymous SNVs,
however, are usually assumed to be neutral. \FIG{aft} shows the time course of
the frequencies of all synonymous and nonsynonymous viral mutations observed in
\env, in \shankaregion, in a chronically HIV-1 infected patient (p10
from \citet{shankarappa_consistent_1999}). Despite many synonymous SNVs
reaching high frequency, very few fix (panel~\ref{fig:aftsyn}); in constrast, many
nonsynonymous mutations fix (panel~\ref{fig:aftnonsyn}). This basic observation
seems at odds with the assumption of neutrality.

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\subsection{Many synonymous polymorphisms in \shankaregion{} are deleterious}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
We study the dynamics and fate of synonymous SNVs more quantitatively by
analyzing data from seven patients from \citet{shankarappa_consistent_1999} and
\citep{liu_selection_2006} as well as three patients from
\citet{bunnik_autologous_2008} (patients with strong viral population structure
were excluded from the analysis; see methods and \figurename~S\PCApat). The
former data set is restricted to the \shankaregion{} region of \env, while the
data from \citet{bunnik_autologous_2008} cover most of \env.  We consider all
SNVs in a frequency interval $[\nu_0-\delta\nu, \nu_0+\delta\nu]$ at some time
$t$, and calculate the fraction that are still observed at later times $t+\Delta
t$. Plotting this fraction against the time interval $\Delta t$, we see that
most synonymous SNVs segregate for roughly one year and are lost much more
frequently than expected under neutrality (panel \ref{fig:fixp1}). The long-time
probability of fixation, $\pfix$, is shown as a function of the initial
frequency $\nu_0$ in panel \ref{fig:fixp2}. We find that $\pfix$ of synonymous
variants is far below the neutral expectation in \shankaregion (red line).
Outside of \shankaregion, using data from \citet{bunnik_autologous_2008} only,
we find no such reduction in $\pfix$ (green line). Restricted to the
\shankaregion{} region, the sequence samples from \citet{bunnik_autologous_2008}
are fully compatible with data from \citet{shankarappa_consistent_1999}. The
nonsynonymous SNVs seem to follow more or less the neutral expectation (blue
line) -- a point to which we will come back below.

\begin{figure}
\begin{center}
\subfloat{\includegraphics[width=\linewidth]
{Shankarappa_allele_freqs_trajectories_syn_p10.pdf}
\label{fig:aftsyn}}\\
 \subfloat{\includegraphics[width=\linewidth]
{Shankarappa_allele_freqs_trajectories_nonsyn_p10.pdf}
\label{fig:aftnonsyn}}
\caption{Time series of frequencies
of synonymous (A) and nonsynonymous (B) single nucleotide variants (SNVs) in \env, 
\shankaregion, from patient p10~\cite{shankarappa_consistent_1999}.
While many nonsynonymous SNVs fix, few synonymous
SNVs do so even though they are frequently observed at high
frequencies. Colors indicate the position of the site along the \shankaregion{} region
(blue to red). Inset: the fixation probability $\pfix$ of a neutral
SNV that reached 50\% frequency is one half.}
\label{fig:aft}
\end{center}
\end{figure}

When interpreting these results for the fixation probabilities, it is important
to note that we focus on SNVs that have already reached high frequencies. In
HIV-1 infection, most SNVs remain very rare throughout; they are not considered
here. Synonymous SNVs can reach high frequencies by two means, either genetic
drift or genetic hitchhiking on escape variants (see below); very deleterious
variants will never reach high frequencies in the first place. Hence, our
analysis indicates that, among all synonymous SNVs that somehow reach high
frequencies, most of them are deleterious in \shankaregion{}, while they tend to
be neutral in the rest of \env{}.

\begin{figure}
\begin{center}
\subfloat{\includegraphics[width=0.9\linewidth]{fixation_times.pdf}
\label{fig:fixp1}}\\
\subfloat{\includegraphics[width=0.9\linewidth]{fixation_probabilities.pdf}
\label{fig:fixp2}}
\caption{Fixation and loss of SNVs.
Panel A) shows how quickly synonymous SNVs are purged from the populations. 
Specifically, the figure shows the fraction of SNVs that are still observed
after $\Delta t$ days, conditional on being observed in one of the three frequency 
intervals (different colors). 
In each frequency interval, the fraction of synonymous
SNVs that ultimately survive is the fixation probability $\pfix$ conditional on the
initial frequency. The neutral expectation for $\pfix=\nu_0$ is indicated by 
dashed horizontal lines.
Panel B) shows the fixation probability of synonymous SNVs as a function of $\nu_0$. Polymorphisms within \shankaregion{} fix less
often than expected for neutral SNVs (indicated by the diagonal line).
This suppression is not observed in other parts of \env{} or for nonsynonymous
SNVs.
The horizontal error bars on the abscissa are bin sizes, the vertical ones the
standard deviation after 100 patient bootstraps of the data. Data from
refs.~\cite{shankarappa_consistent_1999,liu_selection_2006, bunnik_autologous_2008}.}
\label{fig:fixp}
\end{center}
\end{figure}

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\subsection{Synonymous mutations in \shankaregion{} tend to disrupt RNA stems}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
One possible explanation for lack of fixation of synonymous SNVs in
\shankaregion{} is secondary structures in the viral RNA, the disruption of which
is deleterious to the virus \citep{forsdyke_reciprocal_1995,
snoeck_mapping_2011, sanjuan_interplay_2011}.

The propensity of nucleotides in the HIV-1 genome to form base pairs has been
measured using the SHAPE assay, a biochemical reaction preferentially altering
unpaired bases (the HIV-1 genome is a single stranded RNA)
\citep{watts_architecture_2009}. The SHAPE assay has shown that the variable
regions V1-V5 tend to be unpaired, while the conserved regions between those
variable regions form stems. We aligned the within-patient sequence samples to
the reference NL4-3 strain used in \citet{watts_architecture_2009} and assigned
SHAPE reactivities to most positions in the alignment. We then calculated the
distributions of SHAPE reactivities for synonymous SNVs that fixed or were
subsequently lost (only SNVs with frequencies above 15\%). As shown in
\FIG{SHAPEA}, the reactivities of fixed SNVs (red histogram) are systematically
larger than those of lost SNVs (blue) (Kolmogorov-Smirnov test on the cumulative
distribution, $p\approx 0.002$). In other words, SNVs that are likely to
break RNA helices are also more likely to revert and finally be lost from the
population, restoring the helix. Note that this analysis will be sensitive only
at positions where the base pairing pattern of NL4-3 agrees with that of each
patient's initial consensus sequence -- it is thus statistically conservative.
As a control, we also calculate the distribution of SHAPE reactivities for SNVs
that never reach high frequencies (green). This set is a mixture of neutral and
deleterious SNVs and, as expected, their distribution lies between those of
fixed and lost high-frequency SNVs.

To test the hypothesis that synonymous SNVs in \shankaregion{} are lost because they
break stems in the conserved stretches between the V loops, we consider
separately SNVs in V loops and conserved flanks. The greatest
depression in fixation probability is observed in the conserved stems, while the
V loops show little deviation from the neutral signature, see
\FIG{SHAPEB}. This is suggestive of important RNA helices in conserved
regions between the V loops.

In addition to RNA secondary structure, we have considered other possible
explanations for a fitness cost of some synonymous mutations, in particular
codon usage bias (CUB). HIV-1 is known to prefer A-rich codons over highly
expressed human codons \citep{jenkins_extent_2003, kuyl_biased_2012}. We do not
find, however, any evidence for a contribution of average CUB to the ultimate
fate of synonymous SNVs; this agrees with the observation that HIV-1 is not
adapting its codon usage to its human host cells at the macroevolutionary level
\citep{kuyl_biased_2012}.

\begin{figure}
\begin{center}
\subfloat{\includegraphics[height=0.50\linewidth]{reactivities_histograms_syn.pdf}\label{fig:SHAPEA}}
\subfloat{\includegraphics[height=0.49\linewidth]{fixation_probabilities_VnonV.pdf}\label{fig:SHAPEB}}
\caption{Permissible synonymous mutations tend to be unpaired.
Panel A) shows the distribution of SHAPE reactivities among sites at which synonymous 
SNVs fixed (red), sites at which SNVs reached frequencies above 15\% but
were susequently lost (blue), and sites at which no high-frequency SNVs were observed (green) 
(all categories are restricted to the regions V1-V5$\pm 100$bp).
Sites at which SNVs fixed tend to have higher SHAPE reactivities, corresponding to
less base pairing, than those at which SNVs are lost.
Sites at which no SNVs are observed show an intermediate distribution of SHAPE values.
Panel B) shows the fixation probability of synonymous SNVs in
\shankaregion{} separately for variable regions V3-V5 and the connecting conserved 
regions C2-C4 that harbor RNA stems. As expected, the fixation probability is lower
inside the conserved regions. Data from Refs.~\cite{shankarappa_consistent_1999,
bunnik_autologous_2008, liu_selection_2006}.}
\label{fig:SHAPE}
\end{center}
\end{figure}

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\subsection{Deleterious SNVs reach high frequency by hitchhiking}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

While the observation that some fraction of synonymous SNVs is deleterious
is not unexpected, it seems odd that we observe them at high population
frequency and that the fixation probability is reduced only in parts of the
genome (in \shankaregion{} but not in the rest of \env{}; compare the red
triangle line versus the green square line in \FIG{fixp2}).
The region \shankaregion{} undergoes frequent adaptive changes to evade
recognition by neutralizing antibodies \cite{williamson_adaptation_2003,
richman_rapid_2003}. Due to the limited amount of recombination in HIV-1
\cite{neher_recombination_2010, batorsky_estimate_2011}, deleterious SNVs
that are linked to adaptive variants can reach high frequency. This process is
known as hitchhiking \citep{smith_hitch-hiking_1974} or genetic draft
\citep{gillespie_genetic_2000,neher_genetic_2011}. Hitchhiking is apparent in
\FIG{aft}, which shows that many SNVs change rapidly in frequency as a
flock.

The approximate magnitude of the deleterious effects can be estimated from
\FIG{fixp1}, which shows the distribution of times after which synonymous
SNVs at intermediate frequencies become fixed or lost. The typical time to
loss is of the order of 500 days. If this loss is driven by the deleterious
effect of the mutation, this corresponds to deleterious effects $s_d$ of the
order of $- 0.002$ per day. (This is only an average estimate: every single
mutation is expected to have a slightly different fitness effect.)
